R Scripting and Loops
========================================================

# First we will need to set up your working directory
1. Set up working director:
    + Download "Download Data and Scripts", in your download directory, you will see rscripting_workshop.zip, unzip it and copy rscripting_workshop folder into "RNanoCourse"
    + Click Session - Choose Directory -> Desktop -> RNanoCourse -> rscripting_workshop


# Writing your own functions

>The basics of a function are:
 FunctionName <- function(args1,args2, ...) {
    statements
    return (something)
}

1. Writing a simple function. Paste the following code
in your console.
```{r}
SumOfNumbers <- function(x,y) {
  z = x + y
  return (z)
}
```

This is a a simple function called SumOfNumbers,
which takes two arguments x and y and returns
their sum.

2. Use the function SumOfNumbers to add x=35 and y=43?
```{r}

```

3. Create a function SumOfSquares which takes two
arguments x and y and returns the sum of the squares

```{r}

```

4. Functions can return a list of multiple
arguments for more complex calculations,
using the list() function
```{r}
MultipleCalculations <- function(x,y) {
  z = x + y
  c = x^2 + y^2
  a = list(result1=z, result2=c)
  return (a)
}
```

5. Try the function on MultipleCalculations on
x=5 and y=4? What does it return
```{r}

```

6. Functions can take a variety of argument types
* single numbers
* vectors
* matrices

Make the vectors v1 and v2 and use the function
SumOfNumbers to add the two. What is the result?
```{r}
v1 <- c(1,4,5)
v2 <- c(-1,20,3)
SumOfNumbers(v1,v2)
```

Make the matrices m1 and m2 and use the function
SumOfSquares to get the sum of squares the two.
What is the result?
```{r}
m1 <- matrix( c(4, 3, -3, 1, -3, 4), ncol=2)
m2 <- matrix( c(-2, 2, -2, 0 , 5, 2), ncol=2)

```


### Summary:
* Keep your functions short.
* Add comments on what inputs are, what the function does
and what is the output.
* Check for errors along the way.
* Try out your function with simple examples to make sure it’s working properly

# Loops

We have written function that perform a single operation.
But we want to write a function that can be used many times
to do the same operation on lots of different data.

* For: Allows operations to be executed repeatedly
* While: Allows operations to be executed repeatedly until a condition is meet
* Repeat: Allows operations to be executed repeatedly until a a user defined
condition is meet

## For Loops

The for loop is used when iterating through a list.

7. A simple for loop that prints out numbers 1 through 4:
```{r}
for(i in 1:4) {
    print(i)
}
```

8. Create a matrix and find the sum of each row.
```{r}
example.matrix <- matrix( c(4, 8, 31, 1, 0, 20, 10, 20, 15,12, 34, 2), ncol=4)

MatrixRowSum <- function(m){
    vector.row <- matrix()
    for (i in 1:nrow(m)){
        row.sum <- 0
        for (j in 1:ncol(m)){
            row.sum <- row.sum + m[i,j]
        }
        vector.row[i] <- row.sum
    }
    return(vector.row)
}

MatrixRowSum(example.matrix)
```

9. Write the a function, MatrixRowMean to find the mean of each row
for example.matrix.
```{r}

```

## Apply functions
There are several function in R that allow you to apply some function to a
series of objects (eg. vectors, matrices, dataframes or files). They include:

* apply : Applies a function to every column or row of an array and returns the results in an array.
* lapply: Applies a function to elements in a list or a vector and returns the results in a list.
* sapply: Applies a function to elements in a list and returns the results in a vector, matrix or a list.
* tapply: Applies a function to each cell of an array.


> apply(X, MARGIN, FUN, ...)

* X : Array or Matrix
* Margin: Specify which margin we want to apply the function to and which margin
we wish to keep. If the array we are using is a matrix then we can specify
the margin to be either 1 (apply the function to the rows) or
2 (apply the function to the columns).
* FUN: Can be any function that is built in or user defined to be applied to X
* ...: Additional arguments for that will be passed to the function.



10. Example of finding the sum of each row using the apply function.
```{r}
apply(example.matrix,1,sum)
```


11. Find the mean and the median of each row using the apply function.
```{r}

```

# Writing Scripts

An R script is simply a text file containing the same commands that you would
enter on the command line of R.


## Plotting

### Boxplots

A standard way of showing the distribution of the data.

12. Load iris sample data and install necessary packages. Make a boxplot
of Sepal.Length by Species.
```{r, include=FALSE}
data(iris)
install.packages("ggplot2",repos="http://cran.r-project.org")
install.packages("grid",repos="http://cran.r-project.org")
install.packages("gridExtra",repos="http://cran.r-project.org")
library(ggplot2)
library(grid)
library(gridExtra)
```
```{r}
ggplot(data = iris, aes(Species, Sepal.Length)) + geom_boxplot()
```

### Violin plots
A violin plot is more informative than a plain box plot.
In fact while a box plot only shows summary statistics such as
mean/median and interquartile ranges, the violin plot shows the
full distribution of the data.

13. Make a violin plot of Sepal.Length by Species.
```{r}
ggplot(data = iris, aes(Species, Sepal.Length)) +  geom_violin() + geom_jitter(height = 0)
```

Can we generalize a function to make both boxplot and violin plot
side by side?

Open multi_plots.R and lets walk through the function.

14. Plot Sepal.Length and Species using the MultiPlots function.
```{r}
source('multi_plots.R')
MultiPlots(iris, aes(Species, Sepal.Length), "Boxplot and Violin Plot")
```

15. Update the plot to color the different Species. Also
extend the MultiPlots function to take a name for the y axis.
```{r}

MultiPlots <- function(data, mapping, labs, title) {
    library(ggplot2)
    library(grid)
    library(gridExtra)
    p1 = ggplot(data = data, mapping) + geom_boxplot() + labs
    p2 = ggplot(data = data, mapping) +  geom_violin() + geom_jitter(height = 0) + labs
    grid.arrange(p1, p2, ncol=2, top = title)
}

```

* Please fix the following function call:
+ MultiPlots(iris, aes(Species, Sepal.Length, color=??), labs(y="Length"), ??)

16. Extend the MultiPlots function to that you can plots Sepal.Width
as both boxplot and violin plot in the same page.
(Hint: Add a separate mapping argument for the other plots )
```{r}

MultiPlots <- function(data, mapping1, mapping2, title) {
    library(ggplot2)
    library(grid)
    library(gridExtra)
    p1 = ggplot(data = data, mapping1) + geom_boxplot()
    p2 = ggplot(data = data, mapping1) +  geom_violin() + geom_jitter(height = 0)
    p3 = ggplot(data = data, mapping2) + geom_boxplot()
    p4 = ggplot(data = data, mapping2) +  geom_violin() + geom_jitter(height = 0)
    grid.arrange(p1, p2, p3, p4, ncol=2, top = title)
}

```

* Please fix the following function call:
+ MultiPlots(iris, aes(Species, Sepal.Length), ??, ??)


17. Update the MutliPlots function to make side by side plots
of Sepal.Length and Sepal.Width, with boxplot in the middle of the violin plot
to view the summary the data.
(Hint: Change the width of the boxplot to fit inside the violin plot)
```{r}
MultiPlots <- function(data, mapping1, mapping2, title) {
    library(ggplot2)
    library(grid)
    library(gridExtra)
    p1 = ggplot(data = data, mapping1) + geom_violin()  + geom_boxplot(width=.1)
    p2 = ggplot(data = data, mapping2) +  geom_violin() + geom_boxplot(width=.1)
    grid.arrange(p1, p2, ncol=2, top = title)
}
```

* Please fix the following function call:
+ MultiPlots(iris, ??, ??, ??)

### Scripts provide:
* reproducible – share the script and its results
* re-usable – analyze additional input data/files with the same command(s)
* extensible – never start analysis from scratch
* reliable – no danger of missing typing files

## Load own script
Source your own custom script and run a function from it to perform any of the previous operations
```{r}

```

## Command line

Executing R scripts at the command line allows you to:
* Automate your scripts
* Integrate R into pipelines
* Call R from other tools or systems

18. Install package "argparse", open exp_plot.R script file, and 
```{r, include=FALSE}
install.packages("argparse",repos="http://cran.r-project.org")
library(argparse)
```

* Execute the following command in the terminal
+ RScript exp_plot.R 1.25 6


