---
title: "R Scripting and Markdown"
author: "Christopher Bennett"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# R Script:

#### An R script is simply a text file containing the same commands that you would enter on the command line

## Advantages of Scripting:

1. Make reproducible analyses

2. Easy to make changes to analyses

3. Scripts can be shared with collaborators, reviewers, and colleagues

4. Hands off and simple to run


## Elements of R Scripts:

1. **Settings**: 
    + Libraries
    + Input and Output directory

2. **Input**: 
    + Load data to be processes
    + Arguments

3. **Processes**: 
    + Cleaning 
    + Statistical Analyses
    + Plots

4. **Output**: 
    + Figures
    + Tables
    + Data
    + Results

## Download and set R packages
### By Menu in R Studio
* Menu > Tools > Install Packages > Type name in Packages field

* Packages Window > Check packages

### By Command
```{r, eval=FALSE}
install.packages("ggplot2")
```

```{r}
library(ggplot2)

?require()

```

## Set Working Directory
### Selecting a Working Directory by menu in R Studio
* Menu > Session > Set Working Directory

* Environment window > Files > More > Set As Working Directory / Go To Working Directory

### Selecting a Working Directory by command

```{r}
getwd()
setwd("rscripting_workshop/")
list.files()

```

## Data Input: reading data from a file
```{r, eval=FALSE}
?read.table

```

```{r}
setwd("rscripting_workshop/")
tbl <- read.table(file="mtcars.csv", sep=',', header=TRUE)
head(tbl)

```

### Row and column names
```{r}
setwd("rscripting_workshop/")
tbl <- read.table(file="mtcars.csv", sep=',', header=TRUE, row.names=1)
row.names(tbl)
rownames(tbl)

colnames(tbl)
names(tbl)
```

### Slicing Dataframes
```{r}
head(tbl)

```

```{r}
tbl2 <- tbl[1:5,1:5]
head(tbl2)

```

```{r}
tbl3 <- tbl[1:10,]
head(tbl3)

```

```{r}
tbl4 <- tbl[,4:8]
head(tbl4)
```

## Processes
### Programming Functions
* Loops
    + *for* and *while*
* Conditional statements
    + *if* and *else*
* Functions
    + user-defined calculations
    + calling on 3rd party or built-in functions  
    
#### FOR loops
```{r}
Ys <- c("drat", "disp", "hp","wt")
x <- tbl$mpg
par(mfrow=c(2,2))
for (i in 1:4) {
  y <- tbl[, Ys[i]]
  plot(x,y,xlab="mpg",ylab=Ys[i])
}
```

```{r}
tbl$mpgpwt <- 0
rows <- nrow(tbl)
for(i in 1:rows){
  mpg <- tbl$mpg[i]
  wt <- tbl$wt[i]
  tbl$mpgpwt[i] <- mpg/wt
}
head(tbl)

```

#### Nested FOR loops
```{r}
Ys <- c("drat", "disp", "hp")
Xs <- c("mpg", "gear", "carb")
par(mfrow=c(3,3))
for (i in 1:3) {
  x <- tbl[, Xs[i]]
  for (j in 1:3) {
    y <- tbl[,Ys[j]]
    plot(x,y,xlab=Xs[i], ylab=Ys[j])
  }
}
```

#### WHILE loops
```{r}
i <- 0
square <- 0
while (square <88) {
  i <- i+1
  square <- i*i
}
i-1
```
    
#### IF statements
```{r}
x<-2
if (x>0) {
  log.x <- log2(x)
}
log.x
```

#### IF-ELSE statements
```{r}
x <- 0.1
if (x>1) {
  log.x <- log2(x)
} else {
  log.x <- 0
}
log.x
```

```{r}
tbl$goodmile <- NA
rows <- nrow(tbl)
for(i in 1:rows){
  mpg <- tbl$mpg[i]
  if(mpg > 24){
    tbl$goodmile[i] <- 'Yes'
  } else {
    tbl$goodmile[i] <- 'No'
  }
  
}
head(tbl)

```

### Controlling loops
* BREAK
    + In a conditional statement to stop the loop
* NEXT
    + In a conditional statement to skip the analysis for certain rounds of the loop

#### REPEAT and BREAK
Without a "break" conditional repeats are infinite loops!
```{r}
i <- 0
square <- 0
repeat {
  i <- i+1
  square <- i^2
  if (square > 88) {
    break
  }
}
i-1
```

#### NEXT in a FOR loop
```{r}
sum <- 0
num.cars <- 0
for (i in 1:length(tbl$mpg)) {
  if (tbl$cyl[i] > 6) {
    next
  } else {
    sum <- sum + tbl$mpg[i]
    num.cars <- num.cars+1
  }
}
avg.mpg <- sum/num.cars
sum
num.cars
avg.mpg
```

### Useful function: AGGREGATE
```{r}
?aggregate
aggregate(mpg ~ cyl, tbl, mean)
```

```{r, eval=FALSE}
aggregate(. ~ cyl, tbl[,-13], median)
```

```{r}
boxplot(tbl$mpg ~ tbl$cyl)
```

### Useful function: SUMMARY
```{r}
?summary
summary(tbl)
```

### Useful function: REPLICATE
```{r}
replicate(5, seq(1:5))
```

```{r}
replicate(12, rnorm(10))
```

### User defined functions
There are lots of built-in functions in R. But, sometimes, you need to make custom code for a process.

Functions are just a set of instructions(calculations) that we want to use repeatedly or that, because of their complexity, are better self-contained in a sub-program and called when needed.

#### Basic FUNCTION elements
```{r}
function.name <- function(arguments)
{
  #computations on the arguments
  #and some other code
}
```

```{r}
square <- function(x)
{
  x^2
}

square(40)

k <- 10
square(k)
```

#### Setting a default value or range
```{r}
square <- function(x=10, n=seq(0.05, 1, by=0.01))
{
  x^n
}
square(2,5)

square(,2)

square(2,)
```

### Calling functions in a function
```{r}
my.fun <- function(X.matrix, y.vec, z.scalar)
{
  sq.scalar <- square(z.scalar,2)
  mult <- X.matrix %*% y.vec
  final <- mult * sq.scalar
  return(final)
}

my.mat <- matrix(c(1:6),3,2)
my.vec <- c(3:4)

my.fun(my.mat, my.vec, 5)
```

### Function returning a List
```{r}
my.fun <- function(X.matrix, y.vec, z.scalar) 
{
  sq.scalar <- square(z.scalar,2)
  mult <- X.matrix %*% y.vec
  final <- mult * sq.scalar
  return(list(sq.num=sq.scalar, matmult=final))
}

my.fun(my.mat, my.vec, 5)
```
### Function Best Practices
* keep your functions short.

* If things start to get very long, you can probably split up your function into more manageable chunks that call other functions. This makes your code cleaner and easily testable.

* Functions make your code easy to update. You only have to change one function and every other function that uses that function will also be automatically updated.

* Put in comments on what are the inputs to the function, what the function does, and what is the output.

* Check for errors along the way.

* Try out your function with simple examples to make sure it's working properly.

### APPLY functions
* The APPLY family can be used to perform functions to manipulate slices of data from matrices, arrays, lists and data frames in a repetitive way.

* 'apply' - operates on array or matrix

* 'lapply' and 'sapply' - traversing over a set of data like a list or vector, and calling the specified function for each item. sapply return a vector and lapply returns a list

* 'mapply' - multivariate apply.

* 'tapply' - applies a function to each cell of an array

```{r}
# (N)apply(X, MARGIN, FUN, ...)
```

* apply
```{r}
apply(tbl[-13], 2, sum) #sum of each column in tbl

ColMax <- function(x) apply(x,2,max)
ColMax(tbl[-13])
```

* sapply/lapply
```{r}
sapply(1:3, function(x) x^2)
```

```{r}
lapply(1:3, function(x) x^2)
```

* mapply
```{r}
mapply(rep, 1:4, 4:1)
```

* tapply
```{r}
tapply(tbl$mpg, tbl$cyl, mean)
```

#### APPLY for repetitive application 
```{r}
r <- rainbow(3) #rainbow(n): create a vector of n contiguous colors
choose.col <- function(n) {
  if (n>3) {
    col.n <- r[1]
  }
  if (n>5) {
    col.n <- r[2]
  }
  if (n>7) {
    col.n <- r[3]
  }
  col.n
}

col.pch <- sapply(tbl$cyl, choose.col)
cbind(tbl$cyl, col.pch)
```

### Calling functions in Plot
```{r}
plot(tbl$mpg, tbl$hp, col=sapply(tbl$cyl, choose.col))
legend("topright", legend=c(4,6,8), col=r, pch=1)
```

### Calling functions
* Functions can be stored in the script.

* To use functions in many scripts, they can be saver in their own files or as a function set.

* Use SOURCE to call functions in another file:
  + e.g. source("square_functions.R")

## Output:
### Save and Load
* R objects (variables) can be saved into a file:
    + save(mult.list, file='mult_list.Rda')
    
* Saved objects can be loaded into a new session:
    + load('mult_list.Rda')
    
### Export a table
* write.csv()
    + write.csv(mydata, "mydata.txt, quote=TRUE, row.names=TRUE)
    
* write.table()
    + (tab delimited) write.table(mydata, "mydata.txt, sep="\t", quote=FALSE, row.names=TRUE)
    
    + (comma delimited) write.table(mydata, "mydata.txt", sep=",", quote=TRUE, row.names=TRUE)

* write.xlsx()
    + library(xlsx)
    + write.xlsx(mydata, "mydata.xlsx")
    
### Graphical outputs
* postscript()
    + postscript(file="cool_plot.ps", paper="letter", horizontal=TRUE)
    
* png()
    + pgn(filename="mpg_by_cyl.png", width=480, height=480)
    
* tiff()
    + tiff(filename="mpg_by_cyl.tiff", width=480, height=480)
    
## Putting it all together
```{r, eval = FALSE}
r <- rainbow(3) #rainbow(n): create a vector of n contiguous colors
choose.col <- function(n) {
  if (n>3) {
    col.n <- r[1]
  }
  if (n>5) {
    col.n <- r[2]
  }
  if (n>7) {
    col.n <- r[3]
  }
  col.n
}

sep.csv <- c(',') #for tab-delimited# sep.tab <- c('\t')
csv.file <- c("mtcars.csv")
tbl <- read.table(file=csv.file, sep=sep.csv, header=TRUE)

postscript(file="cool_plot.ps", paper="letter", horizontal = TRUE)
plot(tbl$mpg, tbl$hp, col=sapply(tbl$cyl, choose.col))
legend("topright", legend=c(4,6,8), col=r, pch=1)
dev.off()

mpg.cyl <- aggregate(mpg ~ cyl, tbl, mean)
write.table(mpg.cyl, file="mpg_mean_cyl.txt", sep="\t", row.names = FALSE)
png(filename = "mpg_by_cyl.png", width=480, height=480)
boxplot(tbl$mpg ~ tbl$cyl)
dev.off()
```

## Load custom functions as a library
* Allows reuse of custom functions
* Keeps things tidy

```{r, eval = FALSE}
source("your_script.R")

```

## Run script
* In Editor, select lines to run and click 'Run' icon

* Copy lines to run from Editor and paste them into Consol (+ ENTER)

* On terminal, use 'Rscript' command
    + e.g. $ Rscript plot_mpg.R
    
## Command Line Arguments
* commandArgs
    + accepts values on the command-line and pushes them into an array in the order of the values
    
* argparse
    + accepts values on the command-line using "command line options"
    + prints out help messages

## commandArgs
* save this block as "exp_plot.R" and remove the #s
```{r, eval=FALSE}
args <- commandArgs(TRUE) #get variables from command line
num1 <- as.numeric(args[1])
num2 <- as.numeric(args[2])

square <- function (x,n=seq(1, num2, by=1)) {
  x^n
}

x <- c(1:num2)
y <- square(num1)
png(filename = "exp_plot.png", width = 480, height = 480)
plot(x,y, ylab=paste(num1, "^x", sep=""))
dev.off()
```

* On the command line: $ Rscript exp_plot.R 2 10

## argparse
* save this block as "exp_plot2.R" and remove the #s
```{r, eval = FALSE}
library(argparse)
parser <- ArgumentParser()

parser$add_argument("-n", "--number", type="integer", default=2, help="The number that #will be multiplied by itself", metavar="number")

parser$add_argument("-x", "--exponent", type="integer", default=10, help="The number of #the times -n is multiplied itself (exponent)", metavar="number")

args <- parser$parse_args()
num1 <- args$number
num2 <- args$exponent

square <- function (x,n=seq(1, num2, by=1)) {
  x^n
}

x <- c(1:num2)
y <- square(num1)
png(filename = "exp_plot2.png", width = 480, height = 480)
plot(x,y, ylab=paste(num1, "^x", sep=""))
dev.off()
```

# R Markdown

#### An R Markdown is essentially a file combining executable code and plain text with formatting. It's very useful for exploratory analyses and documenting analysis steps and results.

[https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

# 10-Minute Break before Workshop



